---
title: "Statistical inference with the GSS data"
output: 
  html_document: 
    fig_height: 4
    highlight: pygments
    theme: spacelab
---

## Setup

### Load packages

```{r load-packages, message = FALSE}
library(ggplot2)
library(dplyr)
library(statsr)
```

### Load data

Make sure your data and R Markdown files are in the same directory. When loaded
your data file will be called `gss`. Delete this note when before you submit 
your work. 

```{r load-data}
load("gss.Rdata")
```



* * *

## Part 1: Data

Since, randomly sampled respondents took the survey, therefore, we can term the results as generalisable. Causality cannot be inferred due to no random assignment.

* * *

## Part 2: Research question

Does there appear to be a relationship between religious preference and working in military service?
Minorities getting very low chances to serve in military can result in religious descrimination.
Null Hypothesis: religious preference and working in military service are independent.
Alternate Hypothesis: religious preference and working in military service are dependent.

* * *

## Part 3: Exploratory data analysis

```{r}
armed_proportion <- gss %>% filter(!is.na(vetyears)) %>%
  group_by(vetyears) %>%
  summarise(count=n())
armed_proportion
```

More than 50% of the population didn't respond about working in military service. People serving in army for 2 to 4 years are high with 9.7 %, as compared to people working less than 2 years with 3.7 % and people working for more than 4 years with 3.8 %. Majority of the respondents didn't serve the army with 82.6 %.

```{r}
armed_proportion <- armed_proportion %>% mutate(proportion = count/sum(count))
ggplot(data=armed_proportion,aes(x=vetyears,y=count)) + geom_bar(stat="identity")

```

```{r}
religion_proportion <- gss %>% filter(!is.na(relig)) %>%
  group_by(relig) %>%
  summarise(count=n())
religion_proportion <- religion_proportion %>% mutate(proportion = count/sum(count))
religion_proportion
```
The table of religious preferences reflects that Protestant is the largest religion to be preferred by the population with 59 % of the sample population. Catholic with 24.5 % and Jewish with 2 % are preferred next. 

```{r}
ggplot(data=religion_proportion,aes(x=relig,y=count)) + geom_bar(stat="identity")+ theme(axis.text.x=element_text(angle=90, hjust=1))

```

```{r}
rel_vs_military <- table(gss$vetyears, gss$relig)
rel_vs_military

```
The table shows the count for each religion with respect to the time period spent working in military service. Some respondents never worked, some worked for a few time and some worked for more than 4 years. We can clearly see that minority religious individuals haven't worked too much in military. Is it religious discrimination that stops the minorities to serve the American Army?

* * *

## Part 4: Inference

Since, we are working with 2 categorical variables with one atleast having more than 2 levels, therefore, a chi-squared independence test will be used. The observed counts are compared with the expected counts. Chi squared test along with degrees of freedom will be used to calculate p-value to conclude the relationship between two  categorical variables.

The requirement for a chi-square test are as follows:
1.Independence: Since, USA is a large populated country and sample size is less than 10 % of the entire population, we can conclude that samples are independent due to random sampling without replacement. 

2.Sample Size: Each particular cell must have atleast 5 expected cases which isn't true here for all cells. Therefore, we will calculate p-value using simulation.

```{r}
chisq <- chisq.test(rel_vs_military)
round(chisq$expected, 1)
```
```{r}
chisq2 <- chisq.test(rel_vs_military, simulate.p.value= TRUE, B=100000)
round(rel_vs_military - chisq2$expected, 1)
```
```{r}
chisq2$p.value
```

Since, the p-value is greater than 0.05, therefore there is no association between religious preferance and working in military service.
Hypothesis testing and confidence interval cannot be applied to this data set because we are dealing with more than 1 categorical variables with multiple levels.